---
show-solutions: false
---

# Assignment 1 {.unnumbered}

This assignment focuses on the foundations of the linear model: 

+ matrix representation, 
+ least squares estimation, 
+ projection
+ basic inference. 

::: {.callout-note title="Instructions"}

- Submit both your source file (`.qmd` or `.Rmd`) and the rendered HTML or PDF file.
- Name your source file as `A1_LastName_FirstName.qmd` or `A1_LastName_FirstName.Rmd`.
- Clearly label every question and subpart.
- Type all mathematical work using LaTeX.
- Include all R code used to obtain your answers.
- Your document must render from beginning to end without error.
- Do not submit screenshots of code, R output, or mathematical derivations.
- Round numerical answers to three decimal places when appropriate.
- For True/False questions, enter only `T` or `F`.
- For short-answer questions, one or two clear sentences are sufficient unless a derivation is requested.

::: 

The total value of the assignment is [**40**].

## Question 1: Matrix Representation of the Linear Model [8]

Consider the simple linear regression model

$$
Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i,
\qquad i=1,\ldots,5,
$$

with the following observed data:

| $i$ | $x_i$ | $Y_i$ |
|---:|---:|---:|
| 1 | 0 | 2 |
| 2 | 1 | 4 |
| 3 | 2 | 5 |
| 4 | 3 | 8 |
| 5 | 4 | 9 |

### (a) [2]

Which of the following is the correct design matrix $\mathbf{X}$?

- [ ] **A.**

$$
\begin{pmatrix}
0 & 2\\
1 & 4\\
2 & 5\\
3 & 8\\
4 & 9
\end{pmatrix}
$$

- [ ] **B.**

$$
\begin{pmatrix}
1 & 0\\
1 & 1\\
1 & 2\\
1 & 3\\
1 & 4
\end{pmatrix}
$$

- [ ] **C.**

$$
\begin{pmatrix}
0 & 1\\
1 & 1\\
2 & 1\\
3 & 1\\
4 & 1
\end{pmatrix}
$$

- [ ] **D.**

$$
\begin{pmatrix}
1 & 2\\
1 & 4\\
1 & 5\\
1 & 8\\
1 & 9
\end{pmatrix}
$$

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:** B

:::

### (b) [2]

Fill in the following table.

| Quantity | Dimension / Rank |
|:---|:---|
| $\mathbf{Y}$ | `___` |
| $\mathbf{X}$ | `___` |
| $\boldsymbol{\beta}$ | `___` |
| $\operatorname{rank}(\mathbf{X})$ | `___` |

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:**

| Quantity | Dimension / Rank |
|:---|:---|
| $\mathbf{Y}$ | $5\times 1$ |
| $\mathbf{X}$ | $5\times 2$ |
| $\boldsymbol{\beta}$ | $2\times 1$ |
| $\operatorname{rank}(\mathbf{X})$ | $2$ |

:::

### (c) [2]

Which pair is correct?

- [ ] **A.**

$$
\mathbf{X}^\top\mathbf{X}
=
\begin{pmatrix}
5 & 10\\
10 & 30
\end{pmatrix},
\qquad
\mathbf{X}^\top\mathbf{Y}
=
\begin{pmatrix}
28\\
77
\end{pmatrix}
$$

- [ ] **B.**

$$
\mathbf{X}^\top\mathbf{X}
=
\begin{pmatrix}
5 & 5\\
5 & 30
\end{pmatrix},
\qquad
\mathbf{X}^\top\mathbf{Y}
=
\begin{pmatrix}
28\\
70
\end{pmatrix}
$$

- [ ] **C.**

$$
\mathbf{X}^\top\mathbf{X}
=
\begin{pmatrix}
30 & 10\\
10 & 5
\end{pmatrix},
\qquad
\mathbf{X}^\top\mathbf{Y}
=
\begin{pmatrix}
77\\
28
\end{pmatrix}
$$

- [ ] **D.**

$$
\mathbf{X}^\top\mathbf{X}
=
\begin{pmatrix}
5 & 10\\
10 & 20
\end{pmatrix},
\qquad
\mathbf{X}^\top\mathbf{Y}
=
\begin{pmatrix}
28\\
77
\end{pmatrix}
$$

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:** A

:::

### (d) [2]

Assume

$$
\mathbb{E}[\boldsymbol{\varepsilon}] = \mathbf{0},
\qquad
\operatorname{Var}(\boldsymbol{\varepsilon})=\sigma^2\mathbf{I}_5.
$$

For each statement, enter `T` or `F`.

| Statement | T/F |
|:---|:---:|
| $\mathbb{E}[\mathbf{Y}] = \mathbf{X}\boldsymbol{\beta}$ | `___` |
| $\operatorname{Var}(\mathbf{Y}) = \sigma^2\mathbf{I}_5$ | `___` |
| $\mathbb{E}[\mathbf{Y}] = \mathbf{0}$ | `___` |
| The five errors all have variance $\sigma^2$ | `___` |

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:**

| Statement | T/F |
|:---|:---:|
| $\mathbb{E}[\mathbf{Y}] = \mathbf{X}\boldsymbol{\beta}$ | T |
| $\operatorname{Var}(\mathbf{Y}) = \sigma^2\mathbf{I}_5$ | T |
| $\mathbb{E}[\mathbf{Y}] = \mathbf{0}$ | F |
| The five errors all have variance $\sigma^2$ | T |

:::

## Question 2: Least Squares and Projection [12]

Continue using the data and design matrix from Question 1.

### (a) [4]

Starting from

$$
S(\boldsymbol{\beta})
=
(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta})^\top
(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}),
$$

derive the normal equations

$$
\mathbf{X}^\top\mathbf{X}\hat{\boldsymbol{\beta}}
=
\mathbf{X}^\top\mathbf{Y},
$$

and then state the OLS estimator when $\mathbf{X}$ has full column rank.

Keep your derivation concise.

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:**

Expanding,

$$
S(\boldsymbol{\beta})
=
\mathbf{Y}^\top\mathbf{Y}
-
2\boldsymbol{\beta}^\top\mathbf{X}^\top\mathbf{Y}
+
\boldsymbol{\beta}^\top\mathbf{X}^\top\mathbf{X}\boldsymbol{\beta}.
$$

Differentiating,

$$
\frac{\partial S(\boldsymbol{\beta})}{\partial\boldsymbol{\beta}}
=
-2\mathbf{X}^\top\mathbf{Y}
+
2\mathbf{X}^\top\mathbf{X}\boldsymbol{\beta}.
$$

Setting the derivative equal to zero gives

$$
\mathbf{X}^\top\mathbf{X}\hat{\boldsymbol{\beta}}
=
\mathbf{X}^\top\mathbf{Y}.
$$

Hence,

$$
\boxed{
\hat{\boldsymbol{\beta}}
=
(\mathbf{X}^\top\mathbf{X})^{-1}
\mathbf{X}^\top\mathbf{Y}
}.
$$

:::

### (b) [3]

Using the quantities from Question 1, compute

$$
\hat{\boldsymbol{\beta}}
=
\begin{pmatrix}
\hat{\beta}_0\\
\hat{\beta}_1
\end{pmatrix}.
$$

Fill in the blanks:

$$
\hat{\beta}_0 = \underline{\hspace{2cm}},
\qquad
\hat{\beta}_1 = \underline{\hspace{2cm}}.
$$

Hence,

$$
\hat{Y}
=
\underline{\hspace{2cm}}
+
\underline{\hspace{2cm}}x.
$$

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:**

$$
\hat{\beta}_0=1.4,
\qquad
\hat{\beta}_1=2.1,
$$

so

$$
\boxed{\hat{Y}=1.4+2.1x}.
$$

:::

### (c) [2]

For the least squares residual vector

$$
\mathbf{e}
=
\mathbf{Y}-\hat{\mathbf{Y}},
$$

which statement must be true?

- [ ] **A.** $\mathbf{X}\mathbf{e}=\mathbf{0}$

- [ ] **B.** $\mathbf{X}^\top\mathbf{e}=\mathbf{0}$

- [ ] **C.** $\mathbf{e}^\top\mathbf{e}=0$

- [ ] **D.** $\mathbf{H}\mathbf{e}=\mathbf{e}$

In one sentence, explain the geometric meaning of your choice.

**Your explanation:**  
`____________________________________________________________`

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:** B

The residual vector is orthogonal to the column space of $\mathbf{X}$.

:::

### (d) [3]

For the hat matrix

$$
\mathbf{H}
=
\mathbf{X}
(\mathbf{X}^\top\mathbf{X})^{-1}
\mathbf{X}^\top,
$$

enter `T` or `F`.

| Statement | T/F |
|:---|:---:|
| $\mathbf{H}^\top=\mathbf{H}$ | `___` |
| $\mathbf{H}^2=\mathbf{H}$ | `___` |
| $\hat{\mathbf{Y}}=\mathbf{H}\mathbf{Y}$ | `___` |
| $\mathbf{H}$ is an orthogonal projection matrix | `___` |

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:**

| Statement | T/F |
|:---|:---:|
| $\mathbf{H}^\top=\mathbf{H}$ | T |
| $\mathbf{H}^2=\mathbf{H}$ | T |
| $\hat{\mathbf{Y}}=\mathbf{H}\mathbf{Y}$ | T |
| $\mathbf{H}$ is an orthogonal projection matrix | T |

:::

## Question 3: Distribution and Inference [12]

Suppose a multiple linear regression model has $n=20$ observations and $p=3$ regression parameters. Assume

$$
\mathbf{Y}
\sim
N_{20}
\left(
\mathbf{X}\boldsymbol{\beta},
\sigma^2\mathbf{I}_{20}
\right).
$$

Suppose

$$
\hat{\boldsymbol{\beta}}
=
\begin{pmatrix}
2.0\\
1.5\\
-0.5
\end{pmatrix},
$$

$$
(\mathbf{X}^\top\mathbf{X})^{-1}
=
\begin{pmatrix}
0.20 & 0 & 0\\
0 & 0.05 & 0\\
0 & 0 & 0.08
\end{pmatrix},
$$

and

$$
\mathrm{SSE}=34.
$$

Use

$$
t_{0.975,17}=2.110.
$$

### (a) [3]

Choose the correct distributions.

For $\hat{\boldsymbol{\beta}}$:

- [ ] **A.** $N_3\!\left(\mathbf{0},\sigma^2\mathbf{I}_3\right)$

- [ ] **B.** $N_3\!\left(\boldsymbol{\beta},\sigma^2(\mathbf{X}^\top\mathbf{X})^{-1}\right)$

- [ ] **C.** $t_{17}$

- [ ] **D.** $\chi^2_{17}$

For $\mathrm{SSE}/\sigma^2$:

- [ ] **A.** $N(0,1)$

- [ ] **B.** $t_{17}$

- [ ] **C.** $\chi^2_{17}$

- [ ] **D.** $F_{3,17}$

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:** B and C

:::

### (b) [2]

Compute

$$
\hat{\sigma}^2
=
\frac{\mathrm{SSE}}{n-p}.
$$

$$
\hat{\sigma}^2 = \underline{\hspace{2cm}}.
$$

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:**

$$
\boxed{\hat{\sigma}^2=2}.
$$

:::

### (c) [4]

For testing

$$
H_0:\beta_1=0
\qquad\text{versus}\qquad
H_1:\beta_1\neq0,
$$

where $\beta_1$ is the second entry of $\boldsymbol{\beta}$, fill in the following:

$$
\mathrm{SE}(\hat{\beta}_1)
=
\underline{\hspace{2cm}},
$$

$$
t
=
\underline{\hspace{2cm}},
$$

and choose the decision:

- [ ] **A.** Reject $H_0$

- [ ] **B.** Fail to reject $H_0$

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:**

$$
\mathrm{SE}(\hat{\beta}_1)
=
\sqrt{2(0.05)}
=
0.3162,
$$

$$
t
=
\frac{1.5}{0.3162}
=
4.7434.
$$

Since $|4.7434|>2.110$, the answer is **A: Reject $H_0$**.

:::

### (d) [3]

Which of the following is the 95% confidence interval for $\beta_1$?

- [ ] **A.** $(0.8329,\;2.1671)$

- [ ] **B.** $(1.1838,\;1.8162)$

- [ ] **C.** $(-0.6671,\;0.6671)$

- [ ] **D.** $(0,\;3.0000)$

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:** A

:::

## Question 4: Linear Regression in R [8]

For this question, use the built-in R dataset `cars`.

The variables are:

- `speed`: speed of the car in miles per hour;
- `dist`: stopping distance in feet.

Consider the model

$$
\mathrm{dist}_i
=
\beta_0
+
\beta_1\mathrm{speed}_i
+
\varepsilon_i.
$$

Load the data using

```{r}
data(cars)
```

### (a) [4]

Fit the model using `lm()` and complete the table.

```{r}
fit <- lm(dist ~ speed, data = cars)
summary(fit)
```

| Quantity | Your answer |
|:---|:---|
| $\hat{\beta}_0$ | `___` |
| $\hat{\beta}_1$ | `___` |

Which interpretation of $\hat{\beta}_1$ is correct?

- [ ] **A.** For each 1 mph increase in speed, the stopping distance increases by exactly 3.9324 feet for every car.

- [ ] **B.** For each 1 mph increase in speed, the fitted mean stopping distance increases by approximately 3.9324 feet.

- [ ] **C.** The average speed increases by 3.9324 mph for every extra foot of stopping distance.

- [ ] **D.** A car travelling at 0 mph has a stopping distance of 3.9324 feet.

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:**

$$
\hat{\beta}_0=-17.5791,
\qquad
\hat{\beta}_1=3.9324.
$$

The correct interpretation is **B**.

:::

### (b) [4]

Construct $\mathbf{Y}$ and $\mathbf{X}$ manually and reproduce the OLS coefficients using matrix operations.

Complete the following code:

```{r}
Y <- matrix(cars$dist, ncol = 1)

X <- cbind(
  1,
  cars$speed
)

# Replace NULL with your matrix expression
beta_hat <- NULL

beta_hat
```

Then report

$$
\hat{\boldsymbol{\beta}}
=
\begin{pmatrix}
\underline{\hspace{2cm}}\\
\underline{\hspace{2cm}}
\end{pmatrix}.
$$

::: {.solution .content-visible when-meta="show-solutions"}

**Solution:**

```{r}
beta_hat <- solve(t(X) %*% X) %*% t(X) %*% Y
```

and

$$
\boxed{
\hat{\boldsymbol{\beta}}
=
\begin{pmatrix}
-17.5791\\
3.9324
\end{pmatrix}
}.
$$

:::

## Grading Summary

| Question | Topic | Value |
|:---|:---|---:|
| Question 1 | Matrix representation | 8 |
| Question 2 | Least squares and projection | 12 |
| Question 3 | Distribution and inference | 12 |
| Question 4 | Regression in R | 8 |
| **Total** |  | **40** |
